---
title: "`R` Handout Slides: Introduction to `tidyverse`"
date: "2023-08-24"
output: beamer_presentation
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```



## Introducing `tidyverse`

`tidyverse` package gives us tools for data cleaning and processing

  * "Base `R`" (and other packages) offers other tools 
  * We want to use these _specific_ tools for Data Analysis Activities

\bigskip 

Things to remember: 

  * First time you use it on a computer, run `install.packages()`
  * Include `library(tidyverse)` at start of your `R` scripts



## Overview

Slides and handout cover following functions: 

-   `select()` - pick out specific columns or variables from your data
-   `filter()` - pick out specific rows or observations from your data
-   `arrange()` - reorder or sort rows of your data
-   `mutate()` - create new variables as functions of existing variables
-   `summarize()` - creates summary statistics (with `group_by()`)

\medskip 

We'll also talk about piping via `%>%`.  



## Sample Data Set 

```{r}

person.ID <- c(12, 24, 54, 65)
address   <- c("123 Main St", "274 Long St", 
               "789 Right St", "467 Left St")
employed  <- c(TRUE, TRUE, FALSE, TRUE)
wage.inc  <- c(12500, 15750, 0, 14100)

sample.data <- data.frame(person.ID, address, 
                          employed, wage.inc)

sample.data

```



## Select Columns or Variables using `select()`

```{r}
# Select person.ID and address from sample.data, 
# and store as a new data frame named ID.address.data

ID.address.data <- select(sample.data, person.ID, address)

# Print new data frame 

ID.address.data

```



## Select Rows or Observations using `filter()`

```{r}
# Keep observation with address of "123 Main St"

filter(sample.data, address == "123 Main St")

# Keep observations with wages greater than 14,000

filter(sample.data, wage.inc > 14000)

```




## Select Rows or Observations using `filter()` Con't

```{r}

# Keep observations using two criteria using the 
# "&" symbol - this syntax requires that BOTH 
# conditions be true 

filter(sample.data, employed == TRUE & person.ID < 50)

# Alternatively, return rows that match EITHER 
# condition using the | symbol (read as "OR")

filter(sample.data, employed == TRUE | person.ID < 50)

```



## Sorting Data using `arrange()`

```{r}
# Sort sample.data observations by wage.inc 
# from smallest to largest

arrange(sample.data, wage.inc)

# You can sort by multiple variables, and 
# reverse sort using desc() function

```



## Summarizing Data with `summarize()`

```{r}
# Use summarize to return a data frame with 
# average wages. 

summarize(sample.data, mean.wage.inc = mean(wage.inc))

# Handles multiple summary stats easily

summarize(sample.data, 
          median.wage.inc = median(wage.inc), 
          pct.10.wage.inc = quantile(wage.inc, 0.1), 
          pct.90.wage.inc = quantile(wage.inc, 0.9))

```



## Creating New Variables using `mutate()`

```{r}

# Silly formula just to show what mutate can do

mutate(sample.data, 
       new.var = log(person.ID)^2 + wage.inc/50)

```



## Piping with `%>%`

```{r}
# Two ways of doing the same thing - first consider: 

sample.data <- mutate(sample.data, 
                      avg.wage.inc = mean(wage.inc))

# Then compare with code below, which does the same 
# but uses %>% to make things clearer: 

sample.data <- sample.data %>%
  mutate(avg.wage.inc = mean(wage.inc))

```



## Piping with `%>%` Continued

```{r}
# Let's consider another example:

summary.stats.table <- sample.data %>% 
  mutate(new.var = log(person.ID)^2 + wage.inc/50) %>% 
  summarize(avg.new.var = mean(new.var))

```



## Grouping Data with `group_by`

```{r}
# Calculating average wages grouped by employed 
# status using group_by()

group.avg.wages <- sample.data %>% 
  group_by(employed) %>% 
  summarize(mean.wages = mean(wage.inc))

```



## Grouping Data with `group_by()` Continued

```{r}
# Create a new variable using mutate

sample.data <- sample.data %>% 
  group_by(employed) %>% 
  mutate(avg.wage = mean(wage.inc))

```
