---
title: "`R` Handout: Learning to Speak `R`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Terminology

`R` is a statistical programming language. In this class, we'll use `R` with 
`RStudio`, which is an IDE (integrated development environment). If you're 
describing your programming experience to someone, you might say "I've used 
`R`" - `RStudio` is _how_ you've worked with `R`. 

## Base `R` and Packages

When you open `R` + `RStudio` for the fist time, there's lots of functions 
that are included by default (e.g., `mean()`, `lm()`, etc.). These functions
are known as "Base `R`" - its what's included with `R` "out of the box." In 
this class, we'll make use of several _packages_ like `tidyverse`. Packages
are collections of functions that aren't part of base `R`. 

To use packages in `R`, you first need to install them using the `install()`
function. You only need to run this command once on your computer. Each time
you open `R`, however, you'll need to _load_ the packages you want to use, 
using the `library()` function. This is a way of telling `R`, "I want to use 
the functions in this package". 

# Scripts and Running Code in `R`

In this class, we'll general work with `R` scripts - in your file explorer, 
these files will have a `.R` file extension. Scripts are collections of `R`
code. To run code in an `R` script, you can click on the line of code (or 
select it with your mouse) and press `Ctrl + Enter` (or `Apple logo / Cmd key + 
Enter` on Mac). 

If you want to run several lines of code at one time, you can highlight all 
of the lines you want to run, then use the hot keys above. You can also use the
`Run` button at the top of your screen. 

## Scripts and the Command Line

When you open `Rstudio`, one of the four panels you'll see is the `Console`. 
You can run code in the command line by clicking to the right of the `>` 
symbol. This can be a convenient way to check a line of code you're writing, 
or to access help files for a function by typing `?` and then the function
name (e.g., you can type `?mean` to read documentation for the `mean` 
function). 

__AN IMPORTANT NOTE__: Whenever you're doing analysis that you want to be able 
to locate or use later, make sure to write your code in a script file that 
you can save! Commands entered in the console won't be saved across `R` 
sessions (meaning, if you close `R` you'll lose what you've entered there). 

## Comments in your Code

An important part of well-written code is _comments_. You can create a comment
by starting a line with a hash tag - `#`. This tells `R`, "ignore this next 
line." Leaving comments is part of _documenting_ your code - explaining what
you're doing so that other people can understand your analysis and read your
code. 

Take a look at the sample `R` output below: 

```{r, include = TRUE, class.output="bg-warning"}
# This is a comment - R will ignore this line

2 + 2 

```

# Basic Data Structures in `R`

Through most of this class, we'll be working with data sets. Before we jump 
into working with data in `R`, however, let's take a second to unpack how
`R` works with data in general. Let's start by creating an _object_ named 
`a` that stores the number `5` in the code below: 

```{r} 

# Create object named a storing the value 5...

a <- 5

# We can now access the value stored in a in equations, functions or other code

a 

a + 2

```

You can find this new object `a` in the `Environment` pane in `RStudio`. 
This object is also available now by calling `a` in the command line.

## The Assignment Operator 

Notice how we used `<-` above to _assign_ the value of `5` to the object `a`?
In `R`, `<-` is the _assignment operator_. You can read it as synonmous with 
"equals". You can use the hot key combination `Alt + -` to save having to 
type both symbols separately. 

## Object Types in `R`

In `R`, we can create several different kinds of objects or data types using the assignment operator. Some of the most important are listed below: 

-   **Numeric**: stores a single value (like `a` stores `5` above)
-   **Logical**: stores binary values `TRUE` or `FALSE`
-   **Character**: also known as string objects, these stores strings of text 
(like "hello world")

We can use `is` functions in `R` to check whether an object is the type that 
we're expecting. These functions will return either `TRUE` or `FALSE`. Take 
a look at the sample code below:

```{r Data Types, echo = TRUE}

# Let's start by checking  if our a object from above is numeric 

is.numeric(a)

# Now we can create a character object and check it using is.character()

a.string <- "123 Main St"

is.character(a.string)

# When we create a logical object, note that we don't put TRUE in quotes

a.logical <- TRUE

is.logical(a.logical)

```

## Vectors in `R` 

Vectors are the basic "building blocks" of data in `R` - they're collections
of items stored together. In general, these items should generally all share 
one of the basic `R` object types. To create vectors, we'll use the
`c()` function to define the list of items we want to combine, like so: 

```{r}

# Create a vector with 3 elements - note the commas between each item!

b.vector <- c(1, 2, 3)

b.vector

# Now we can call functions like mean() using this new vector 

mean(b.vector)

```

Just like our `a` object, `b.vector` is now stored in the `Environment` pane. 

## Dataframes in Base `R`

In Base `R`, data sets are stored as _data frames_ (this will differ slightly
when we start using `tidyverse` but the differences aren't all that important 
right now). Data frames are collections of vectors of the same length. Let's
create a sample data set below: 

```{r}

# To start, let's create vectors storing various object types 

person.ID <- c(12, 24, 54, 65)
address   <- c("123 Main St", "274 Long St", "789 Right St", "467 Left St")
employed  <- c(TRUE, TRUE, FALSE, TRUE)
wage.inc  <- c(12500, 15750, 0, 14100)

# Combine each of the individual vectors into a data frame

data <- data.frame(person.ID, address, employed, wage.inc)

head(data)

```

\newpage 

# List of Useful Functions in Base `R`

Here's a list of commonly used base `R` functions for your reference. Examples 
are included in the code section below. 

Given a vector of numeric values, you can calculate the following: 

  * `mean()`: average
  * `median()`: median
  * `var()`: variance
  * `sd()`: standard deviation
  * `quantile()`: percentiles (requires second argument, see below)
  
You can generate normally-distributed random variables using `rnorm()`, and 
uniformly-distributed random variables using `runif()`. 

```{r}
# Start by creating a vector with 100 draws from a normally-distributed random 
# variables using the rnorm function - set mean = 1 and sd 2

normal.vector <- rnorm(100, mean = 1, sd = 2)

# Now we can use summary stats function above to inspect this vector

mean(normal.vector)
median(normal.vector)
var(normal.vector)
sd(normal.vector)

# We can calculate percentiles using the quantile function and inputting 
# values between 0 and 1 for the percentile we want

quantile(normal.vector, 0.1)  # 10th percentile
quantile(normal.vector, 0.9)  # 90th percentile

```


