w---
title: "`R` Overview: Learning to Speak `R`"
date: "2023-08-24"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Basic Terminology

Two programs we'll use: 

  * `R` is a statistical programming language 
  * `RStudio` is an IDE (integrated development environment)
  
`RStudio` is _how_ we'll work with `R`. 



## Base `R` and Packages

`R` has lots of functions included by default 

  * e.g., `mean()`, `lm()`, etc. 
  * These functions are known as "Base `R`" 

In this class, we'll use several _packages_ like `tidyverse`

  * Packages are collections of functions that aren't in base `R`

Using packages in `R`

  * Install them using the `install()` function (just once)
  * _Load_ them using the `library()` function. 
  
This tells `R`, "I want to use the functions in this package." 



## Scripts and Running Code in `R`

In this class, we'll general work with `R` scripts 

  * These files will have a `.R` file extension. 
  * Scripts are collections of `R` code. 
  
To run code in an `R` script: 

  * Click on the line of code (or select it with your mouse) 
  * `Ctrl + Enter` (or `Apple logo / Cmd key + Enter` on Mac)

If you want to run several lines of code at one time, 
  
  * Highlight all of the lines you want to run, then run 
  * You can also use the `Run` button 
  
  

## Scripts and the Command Line

The `Console` is one of the four default panels in `Rstudio`

  * You can run code by clicking next to the `>` symbol
  * Convenient way to test a line of code
  * Use it to access documentation using "`?`" - e.g, `?mean` 
  
\bigskip

__AN IMPORTANT NOTE__

  * When something's important, make sure its saved in a script!
  * _Don't_ rely on the command line (you'll lose your work)



## Comments in your Code

An important part of well-written code is _comments_. 

  * Start a line with a hash tag - `#`. 
  * This tells `R`, "ignore this next line." 
  
_Documenting_ your code means explaining what you're doing 

  * Helps other people (and you, later!) understand your code
  
\bigskip

```{r, include = TRUE, class.output="bg-warning"}
# This is a comment - R will ignore this line

2 + 2 

```



## Basic Data Structures in `R` and the Assignment Operator

`<-` is the _assignment operator_ - use it to create objects

  * Type it using the hot key combination `Alt + -` 
  * Example code below:
  
\bigskip
  
```{r} 

# Create object named a storing the value 5...

a <- 5

# Now we can use a in equations, functions, etc.

a + 2

```



## Object Types in `R`

In `R`, we can create different _types_ of objects, including: 

-   **Numeric**: stores a single value (like `a` stores `5` above)
-   **Logical**: stores binary values `TRUE` or `FALSE`
-   **Character**: also known as string objects, these stores strings of text 
(like "hello world")

\bigskip

Given an object, we can use "`is`" functions to check types 

  * These functions will return either `TRUE` or `FALSE`
  * Example on next slide



## Object Types in `R` and the `is` Function


```{r Data Types, echo = TRUE}

# Check if a object from above is numeric 

is.numeric(a)

# Now we can create a character object and check it

a.string <- "123 Main St"

is.character(a.string)

```



## Vectors in `R` 

Vectors are the basic "building blocks" of data in `R`

  * Vectors are collections of items stored together
  * To create vectors, we'll use the `c()` function 

\bigskip 

```{r}

# Create a vector with 3 elements - note the commas!

b.vector <- c(1, 2, 3)

b.vector

```



## Dataframes in Base `R`

In Base `R`, data sets are stored as _data frames_ 

  * Data frames are _collections_ of vectors

\bigskip 

```{r}

# Start with vectors storing various object types 

person.ID <- c(12, 24, 54, 65)
address   <- c("123 Main St", "274 Long St", 
               "789 Right St", "467 Left St")
employed  <- c(TRUE, TRUE, FALSE, TRUE)
wage.inc  <- c(12500, 15750, 0, 14100)

```



## Dataframes in Base `R`

```{r}

# Combine individual vectors into a data frame

data <- data.frame(person.ID, address, 
                   employed, wage.inc)

head(data)

```   



